---
title: "Assignment 4"
author: "Kumar Kanishk Singh"
date: "2024-03-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

We fit the given model using JAGS.

```{r}
n<-6
Y<-c(64, 13, 33, 18, 30, 20)
t <- 1:n

alpha<-dnorm(0,10000)
beta<-dnorm(0,10000)
lambda<-exp(alpha+beta*t)
data<-list(Y=Y,n=n)

library(rjags)


model_string<-textConnection("model{

 #Likelihood
 for(t in 1:n){
 Y[t]~dpois(lambda[t])
 }
 
 #Priors
 alpha~dnorm(0,10000)
 beta~dnorm(0,10000)
 for(t in 1:n){
 lambda[t]=exp(alpha+beta*t)
 }
}")

#inits <- list(alpha=alpha, beta=beta)
model <- jags.model(model_string, data = data, quiet = TRUE, n.chains = 2) 

update(model, 2000, progress.bar = "none")

params <-c("alpha","beta")
samples <- coda.samples(model,
                        variable.names = params,
                        n.iter = 5000, progress.bar = "none")

plot(samples)
summary(samples)
```
The trace plot indictes convergence. Now lets look at autocorrelation plots. The autocorrelation is low after lag 2. This also indicates convergence.


```{r}
library(coda)

# Low autocorrelation indicates convergence
autocorr.plot(samples)


autocorr(samples, lag = 1)
```
High ESS indicates convergence
```{r}
effectiveSize(samples)
```
R less than 1.1 indicates convergence
```{r}
gelman.diag(samples)
```
/z/ less than 2 indicates convergence
```{r]
geweke.diag(samples[[1]])
```
We can also see that the rate of discovery increases with time.
```{r}
t<-1:n

a<-summary(samples)$statistics[1,1]
b<-summary(samples)$statistics[2,1]

lambda2<-exp(a+b*t)

plot(lambda2)
```
Lets set up a metropolis sampler for this problem.
```{r}
n<-6
Y<-c(64, 13, 33, 18, 30, 20)
t <- 1:n

beta<-c(3,0)

iter<-25000
par<-matrix(0,iter,2)

cand<-c(0.4,0.06)

posterior<-function(Y,t,beta,sd=10){
  l<-exp(beta[1]+t*beta[2])
  lambda<-prod(dpois(Y,l))
  p<-prod(dnorm(beta,0,sd))
  return(lambda*p)}
#can<-rnorm(1,beta[1],cand[1])
#posterior(Y,t,can)

for(i in 1:iter)
  {
  for(j in 1:2)
  {
    can<-beta
    can[j]<-rnorm(1,beta[j],cand[j])
    R<-posterior(Y,t,can)/posterior(Y,t,beta)
    
    if(runif(1)<R)
    {
      beta<-can
    }
  }
    
  par[i,]<-beta
}

plot(par[,1],type="l",ylab=expression(aplha),xlab="Iteration")
plot(par[,2],type="l",ylab=expression(beta),xlab="Iteration")

```
The trace plots indicate that the model has converged. Following are the acceptance ratios:

```{r}
acc_rate <- colMeans(par[-1,]!=par[-iter,])
acc_rate
```

## Question 2
```{r}
set.seed(27695)
theta_true <- 4
n          <- 30
B          <- rbinom(n,1,0.5)
Y          <- rnorm(n,B*theta_true,1)
hist(Y,breaks=25)
```

b. Following are the plots for f(Y|theta) for theta={2,4,6}:
```{r}
dens<-function(y)
{
  return(exp(-(y^2)/2)/sqrt(2*pi))
}
func<-function(y,theta)
{
  return(dens(y)+dens(y-theta))
}

y<-seq(-3,10,0.1)

plot(func(y,2),type="l",ylab="PDF")
lines(func(y,4),col=2)
lines(func(y,6),col=3)
legend("topright",c("theta=2","theta=4","theta=6"),lty=1,col=1:3,bty="n")
```
c. Following are the MAP estimate of theta and its standard deviation:

```{r stats4}
#install.packages("stats4")
library(stats4)
nlp <- function(theta,Y){
  like<- 0.5*dnorm(Y,0,1) + 0.5*dnorm(Y,theta,1)
  prior<- dnorm(theta,0,10)
  neg_log_post <--sum(log(like))- log(prior)
  return(neg_log_post)}
#map_est <- mle(nlp,start=list(theta=1), fixed=list(Y=Y))
#sd<- sqrt(vcov(map_est))
out <- suppressWarnings(optim(par = 1, nlp, Y = Y, hessian = T))

map_est <- out$par
sd <- sqrt(1 /c(out$hessian))
map_est; sd
```
d. Following are the posterior densities of theta:
```{r}
posterior <- function(theta,Y,k){
  post <- dnorm(theta,0,sqrt(10^k))
  for(i in 1:length(Y)){
    post<-post*(0.5*dnorm(Y[i],0,1)+
                  0.5*dnorm(Y[i],theta,1))
  }
  return(post/sum(post))}

theta <- seq(2,6,0.01)
map <- dnorm(theta,map_est,sd)

plot(theta,map/sum(map),col=1,type="l",ylab="Posterior")
lines(theta,posterior(theta,Y,0),col=2)
lines(theta,posterior(theta,Y,1),col=3)
lines(theta,posterior(theta,Y,2),col=4)
lines(theta,posterior(theta,Y,3),col=5)
legend("topright",c("MAP","k=0","k=1","k=2","k=3"),
       col=1:5,lty=1,bty="n")
```
e.Now we use JAGS to fit this model:
```{r}
data<-list(Y=Y,n=n)

library(rjags)


model_string<-textConnection("model{

 #Likelihood
 for(i in 1:n){
 Y[i]~dnorm(B[i]*theta,1)
 }
 
 #Priors
 for(i in 1:n){
 B[i]~dbern(0.5)
 }
 theta~dnorm(0,1e-2)
}")

#inits <- list(alpha=alpha, beta=beta)
model <- jags.model(model_string, data = data, quiet = TRUE, n.chains = 2) 

update(model, 2000, progress.bar = "none")

params <-c("theta")
samples <- coda.samples(model,
                        variable.names = params,
                        n.iter = 5000, progress.bar = "none")

plot(samples)
summary(samples)
```
We can see that in both part d and e the mean is around 4 and the height of the curve is slightly less than the 0.012.

## Question 3

a.Convergence could be slow because there are more parameters than observations and so it is likely that not all parameters are identifiable.

b. We set up a model using JAGS with lambda = 10 and a = b = 1.

```{r}
Y<-10

data<-list(Y=Y)


library(rjags)


model_string<-textConnection("model{

 #Likelihood
 Y~dbin(p,n)
 
 #Priors
 lambda<-10
  a<-1
  b<-1
 n~dpois(lambda)
 p~dbeta(a,b)
 np<-n*p
}")

#inits <- list(n=20, p=0.5)
model <- jags.model(model_string, data = data, quiet = TRUE, n.chains = 2) 

update(model, 2000, progress.bar = "none")

params <-c("n","p","np")

samples <- coda.samples(model,
                          variable.names = params,
                          n.iter = 5000, progress.bar = "none")
samples_n <- coda.samples(model,
                        variable.names = "n",
                        n.iter = 5000, progress.bar = "none")
samples_p <- coda.samples(model,
                          variable.names = "p",
                          n.iter = 5000, progress.bar = "none")
samples_np <- coda.samples(model,
                          variable.names = "np",
                          n.iter = 5000, progress.bar = "none")

plot(samples_n)
plot(samples_p)
plot(samples_np)
summary(samples)

```

```{r}
library(coda)

# Low autocorrelation indicates convergence
autocorr.plot(samples)


autocorr(samples, lag = 1)

# high ESS indicates convergence
effectiveSize(samples)

# R less than 1.1 indicates convergence
gelman.diag(samples)

# /z/ less than 2 indicates convergence
geweke.diag(samples[[1]])
```
We can see the model converges by the above results. n converges at about 12, p at 0.8 and np at 10.
```{r}
Y<-10

data<-list(Y=Y)


library(rjags)


model_string<-textConnection("model{

 #Likelihood
 Y~dbin(p,n)
 
 #Priors
 lambda<-10
  a<-10
  b<-10
 n~dpois(lambda)
 p~dbeta(a,b)
 np<-n*p
}")

#inits <- list(n=20, p=0.5)
model <- jags.model(model_string, data = data, quiet = TRUE, n.chains = 2) 

update(model, 2000, progress.bar = "none")

params <-c("n","p","np")

samples <- coda.samples(model,
                          variable.names = params,
                          n.iter = 5000, progress.bar = "none")
samples_n <- coda.samples(model,
                        variable.names = "n",
                        n.iter = 5000, progress.bar = "none")
samples_p <- coda.samples(model,
                          variable.names = "p",
                          n.iter = 5000, progress.bar = "none")
samples_np <- coda.samples(model,
                          variable.names = "np",
                          n.iter = 5000, progress.bar = "none")

plot(samples_n)
plot(samples_p)
plot(samples_np)
summary(samples)

```
```{r}
library(coda)

# Low autocorrelation indicates convergence
autocorr.plot(samples)


autocorr(samples, lag = 1)

# high ESS indicates convergence
effectiveSize(samples)

# R less than 1.1 indicates convergence
gelman.diag(samples)

# /z/ less than 2 indicates convergence
geweke.diag(samples[[1]])
```
We can see the model converges by the above results. n converges at about 14, p at 0.6 and np at 8.

## Question 4

We set up a two sample t-test to test whether the treatment is effective or not. We take Yi~Normal(mu,sigma^2) for placebo observations and Yi~Normal(mu+delta,sigma^2) for treatment observations. We test Whether delta=0 or not.
```{r}
n<-12
n1<-6
n2<-6
Y1<-c(2.0,-3.1,-1.0,0.2,0.3,0.4)
Y2<-c(-3.5,-1.6,-4.6,-0.9,-5.1,0.1)

Y1_m<-mean(Y1)
Y2_m<-mean(Y2)

sigma1<-mean((Y1_m-Y1)^2)
sigma2<-mean((Y2_m-Y2)^2)
sig_rt<-sqrt(sigma1/2+sigma2/2)

p_mean<-Y2_m-Y1_m
sigma_hat<-(sig_rt)*sqrt(1/n1+1/n2)


cred_set <- p_mean+sigma_hat*qt(c(0.025,0.975),df=n1+n2)


p_mean;sigma_hat;cred_set
```
The 95% credible set doesn't include 0. So, we can say that the treatment is effective. To test for sensitivity to the prior we also fit the model using proper priors using JAGS.
```{r}
library(rjags)
data <- list(n=6,Y1=Y1,Y2=Y2)

model_string <- textConnection("model{

   # Likelihood
   for(i in 1:n){
     Y1[i] ~ dnorm(mu,tau)
     Y2[i] ~ dnorm(mu+delta,tau)
   }

   # Priors
   mu    ~  dnorm(0, 0.0001)
   delta ~  dnorm(0, 0.0001)
   tau   ~  dgamma(0.1, 0.1)
   sigma <- 1/sqrt(tau)
 }")

model <- jags.model(model_string,data = data, n.chains=2,quiet=TRUE)
update(model, 10000, progress.bar="none")
params  <- c("delta")
samples <- coda.samples(model, 
                        variable.names=params, 
                        n.iter=50000, progress.bar="none")
plot(samples)
summary(samples)
```
We can see that the 95% credible interval now includes 0. So, we can say that the conclusion is slightly sensitive to prior.

## Question 5
```{r}
library(MASS)

# Load the Boston dataset from the MASS library
data(Boston)
n <- nrow(Boston)
p <- ncol(Boston)
X <- cbind(1, Boston[, -p])

# Load necessary libraries
library(rjags)
library(coda)

# Define the data for the model
data_list <- list(Y = Boston$medv, n = n, p = p, X = X)

# Define the JAGS model for part a
model_string_a <- "
model{
  # Likelihood
  for(i in 1:n){
    Y[i] ~ dnorm(inprod(X[i, ], beta[1:p]), tau)
  }

  # Priors
  beta[1] ~ dnorm(0,1e-4)
  for(i in 2:p){
    beta[i] ~ dnorm(0, 1e-4)
  }
  tau ~ dgamma(0.1, 0.1)
  sigma2 <- 1 / tau
}"

# Compile the JAGS model
model_a <- jags.model(textConnection(model_string_a), data = data_list, quiet = TRUE)

# Burn-in
update(model_a, 5e3, progress.bar = 'none')

# Sampling from the posterior distribution
params <- c('beta')
samples_a <- coda.samples(model_a,
                           variable.names = params,
                           n.iter = 2e4, progress.bar = 'none')

# Summarize the posterior samples
summary(samples_a)

# Plot trace and density for beta[1] and beta[2]
par(mfrow=c(2,1))
plot(samples_a[[1]][, "beta[1]"], type="l", main="Trace Plot for beta[1]")
plot(samples_a[[1]][, "beta[2]"], type="l", main="Trace Plot for beta[2]")
par(mfrow=c(2,1))
densplot(samples_a[[1]][, "beta[1]"], main="Density Plot for beta[1]")
densplot(samples_a[[1]][, "beta[2]"], main="Density Plot for beta[2]")


```
```{r}
# Part b - Fit linear regression using lm
fit_least_sq <- lm(medv~., data = Boston)
summary(fit_least_sq)
```
```{r}
# Part c - Define the JAGS model for part c
model_string_c <- "
model{
    # Likelihood
    for(i in 1:n){
      Y[i] ~ dnorm(inprod(X[i, ], beta[]), tau)
    }

    # Priors
    beta[1] ~ ddexp(0,1e-4)
    for(i in 2:p){
      beta[i] ~ ddexp(0, 1e-4)
    }
    tau ~ dgamma(0.1, 0.1)
    sigma2 <- 1 / tau
}"

# Compile the JAGS model for part c
model_c <- jags.model(textConnection(model_string_c), data = data_list, quiet = TRUE)

# Burn-in
update(model_c, 5e3, progress.bar = 'none')

# Sampling from the posterior distribution for part c
samples_c <- coda.samples(model_c,
                           variable.names = params,
                           n.iter = 2e4, progress.bar = 'none')

# Summarize the posterior samples for part c
summary(samples_c)
```
